---
title: My Dissertation
author: Gabriel Voelcker
date: '2018-08-22'
slug: master
categories: []
tags: []
Description: ''
Tags: []
---

This last Monday I presented my dissertation and had the pleasure of concluding my masters. The work is entitled: "An Analysis of Risk-Taking in Family Firms", and was oriented by Prof. PhD. Clea B. Macagnan. 

The main objective of the research is to empirically test whether family firms are more risk averse than non-family firms. I used a sample of brazilian firms, with data from 2010 to 2017, obtaining 1188 observations. I can divide the empirical procedures in two stages. First, I performed tests to compare the average of variables differentiating among family firms and non family firms(62.5% of the tests indicated that they behave differently). Then, I tested my hypothesis using OLS and Random Effects. The results point to a more conservative behaviour from family firms towards risk-taking.

I'd like to acknowledge the contributions of the examination board, which consisted of Profs. PhDs Guilherme Kirch, Cristiano Costa, Roberto Decourt and Daniel Vancin. There is still a lot to work on until this project is published but I'm happy to share for Portuguese speakers the current version of the paper [**here**](https://gabrielvoelcker.netlify.com/FF_Semead.pdf). 



```{r,echo=FALSE,include=FALSE}
library(knitr)    
library(ggplot2)  
library(png)      
```

```{r,echo=FALSE,include=FALSE}
setwd("~/Dropbox/R/Site-Gabriel/content/post")
img1_path <- "~/Dropbox/R/Site-Gabriel/content/post/masters1.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
attr(img1, "info")
```


```{r ,echo=FALSE}
library(png)
library(grid)
setwd("~/Dropbox/R/Site-Gabriel/content/post")
img <- readPNG("masters1.png")
 grid.raster(img)
```






