---
title: My Master's Dissertation
author: Gabriel Voelcker
date: '2018-08-22'
slug: master
categories: []
tags: []
Description: ''
Tags: []
---

This last Monday I presented my dissertation and had the pleasure of concluding an important step: my masters in accounting and finance. The work is entitled: "An Analysis of Risk-Taking in Family Firms", and I was oriented by Prof. PhD. Clea B. Macagnan. 

The main objective of the research is to empirically test whether family firms are more risk averse than non-family firms. The empirical procedures can be divided into two stages. First, I performed tests to compare the average of variables differentiating among family firms and non family firms(62.5% of the tests indicated that they behave differently). Then, I tested my hypothesis using OLS and Random Effects. The results point to a more conservative behaviour from family firms towards risk-taking.

I used a sample of Brazilian firms, with data ranging from 2010 to 2017, totalling 1188 observations. There were two main sources of information that I used. The first is [**Economática**](https://economatica.com/), which is the closest Brazilian equivalent to Compustat. Almost all variables were downloaded from this software, available to all Unisinos students. But I also needed all the information about shareholders, family ties and board of directors, which, by law, must be disclosed by all companies listed in the Brazilian stock market. 

There are two ways of obtaining this information. The first is the more troublesome one: downloading all the available data at the [**source**](https://cvmweb.cvm.gov.br/SWB/Sistemas/SCW/CPublica/CiaAb/FormBuscaCiaAb.aspx?TipoConsult=c). In this case, one would have to download manually for each year and each company of the desired sample.

Luckily, a RStudio package has been developed that quickly downloads all the available data: GetDFPData. It is possible to obtain all the data via RStudio or using the developer's [**website**](http://www.msperlin.com/shiny/GetDFPData/). To anyone interested in obtaining data from Brazilian companies, this is the best free source of data available, in my opinion. All the information provided has already been translated to English. To access the documentation on the package, just click [**here**](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3128252).(I am currently planning to write a post about the relevance and particularities of the Brazilian stock market as well as methods of obtaining its information using RStudio)

I'd like to acknowledge the contributions of the examination board, which consisted of Profs. PhDs Guilherme Kirch, Cristiano Costa, Roberto Decourt and Daniel Vancin. There is still a lot to work on until this project is published but I'm happy to share for Portuguese speakers the current version of the paper [**here**](https://gabrielvoelcker.netlify.com/FF_Semead.pdf). This version has been approved to the [**Semead Congress **](https://www.youtube.com/watch?time_continue=1&v=o9iAqa4wTTM).

```{r,echo=FALSE,include=FALSE}
library(knitr)    
library(ggplot2)  
library(png)      
```

```{r,echo=FALSE,include=FALSE}
setwd("~/Dropbox/R/Site-Gabriel/content/post")
img1_path <- "~/Dropbox/R/Site-Gabriel/content/post/masters1.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
attr(img1, "info")
```


```{r ,echo=FALSE}
library(png)
library(grid)
setwd("~/Dropbox/R/Site-Gabriel/content/post")
img <- readPNG("masters1.png")
 grid.raster(img)
```
