---
output: html_document
---

# 14.121 - Microeconomic Theory I{.tabset}

## **1) Utility and Revealed Preferences**

Based on [**this**](https://gabrielvoelcker.netlify.com/mit/14.121lecture1.pdf).

### Utility Theory

The first class of the semester covers basic Microeconomic topics such as Utility Theory and Revealed preferences. We use an axiomatic approach to the utility theory to find less obvious asumptions of utility theory. It was recognized that utility could not be measured or observed directly, so instead economists devised a way to infer underlying relative utilities from observed choice. 

#### [**Utility Theory**](https://en.wikipedia.org/wiki/Utility): 

In Economics, the concept of **utility** is used to model worth/value. In most economic models, agents are assumed to have a utility function that they maximize. The utility function approach usually serves [**two purposes**](https://www.investopedia.com/ask/answers/12/difference-between-positive-normative-economics.asp):

1) Model for predicting behavior (positive); i.e. agent will maximize utility. This is the "what is" approach. This is most commonly used (95%).
    
2) Framework for welfare analysis (normative); i.e. utility is "good". This is the "what should happen" approach.

If we reflect about this utility approach, we have an opportunity to introduce one useful economic methodology: the **axiomatic approach**.
    
#### [**Axiomatic Approach**](https://en.wikipedia.org/wiki/Axiomatic_system): 

The idea is to *find simple axioms which together are equivalent to
less obvious assumptions of utility theory*.  Any set of axioms from which some or all axioms can be used in conjunction to logically derive theorems exemplify an axiomatic approach. A theory consists of an axiomatic system and all its derived theorems. An axiomatic system that is completely described is a special kind of formal system. A formal theory typically means an axiomatic system, for example formulated within model theory. A formal proof is a complete rendition of a mathematical proof within a formal system.

#### [**Utility Function**](https://en.wikipedia.org/wiki/Utility#Utility_function):

A utility function represents a consumer's preference ordering over a choice set.  Consider a set of alternatives facing an individual, and over which the individual has a preference ordering. A **utility function** is able to represent those preferences if it is *possible to assign a real number to each alternative*, in such a way that alternative a is assigned a number greater than alternative b if, and only if, the individual prefers alternative a to alternative b. In this situation an individual that selects the most preferred alternative available is necessarily also selecting the alternative that maximises the associated utility function. In general economic terms, a utility function measures preferences concerning a set of goods and services. Often, utility is correlated with words such as happiness, satisfaction, and welfare, and these are hard to measure mathematically. Thus, economists utilize consumption baskets of preferences in order to measure these abstract, non quantifiable ideas. 

Gérard Debreu precisely defined the conditions required for a preference ordering to be representable by a utility function. For a finite set of alternatives these require only that the preference ordering is complete (so the individual is able to determine which of any two alternatives is preferred, or that they are equally preferred), and that the preference order is transitive.

##### **Utility Function Definition**: 
A **utility function** u defined on X is a real-valued function u: X -> $\mathbb{R}$. A more foundational place to start to think about assumptions inherent in utility theory is treating preferences as a primitive.

##### **Preference Definition**: 

*P* is a subset of X x X, where we write x $\succeq_P$ *y* if (x,y) $\in$ *P*. Interpretation of x $\succeq_P$ y: agent thinks "*x* is **at least as good as** *y*".

So the utility function *u*: X -> $\mathbb{R}$ **represents** the preference P if:

  u(x) ≥ u(y) $\iff$ x $\succeq_P$ y.

Note: many different utility functions can represent the same set of
preferences. e.g., if X = {x, y, z}, and P = {(x, y),(x, z),(y, z)}
(i.e x $\succeq_P$ y, x $\succeq_P$ z, y $\succeq_P$ z).

In general if u represents P, then so does *f* $\circ$ *u*, for any *strictly* increasing function *f*.

##### **Properties of Preferences**:

- The preference relation $\succeq_P$ is **complete** on *X* if $\forall$*x*, *y* $\in$ *X*, either *x* $\succeq_P$ *y* or *y* $\succeq_P$ *x*.

- The preference relation $\succeq_P$ is **reflexive** on *X* if $\forall$*x* $\in$ *X*, *x* $\succeq_P$ *x*.

- The preference relation $\succeq_P$ is **transitive** on *X* if *x* $\succeq_P$ *y* and *y* $\succeq_P$ *z* $\implies$ *x* $\succeq_P$ *z*.

Our first result is the following **proposition**: if $\succeq_P$ is represented by *u* on *X*, then $\succeq_P$ is complete, reflexive and transitive.

**Proof**:

- Complete: given any *x,y* $\in$ *X*, either u(x) ≥ u(y) or u(y) ≥ u(x) (property of real numbers). This implies x $\succeq_P$ y or y $\succeq_P$ x(by representation assumption). Indifference between two goods doesn't violate completeness.

- Reflexive: given any x $\in$ X, u(x) ≥ u(x) $\implies$ x $\succeq_P$ x.

- Transitive: if x $\succeq_P$ y and y $\succeq_P$ z, then u(x) ≥ u(y) and u(y) ≥ u(z). Transitivity of ≥ on $\mathbb{R}$ implies that u(x) ≥ u(z) $\implies$ x $\succeq_P$ z.

Usually we leave out preferences that do not satisfy these 3 assumptions. Now let's consider the other way around: if $\succeq_P$ is complete, reflexive and transitive, then can it be described by some utility function? It depends on whether the X is **finite**. 

**Proposition**:

Let X be a finite set. Let P be a preference on X. Assume $\succeq_p$ is complete on X, reflexive on X, and transitive on X. Then there exists a utility function *u* which represents P.

**Proof by construction**: B(x) {z $\varepsilon$ X | x $\succeq_p$ z} $\implies$ z is "below" x. Let u(x) = |B(x)| -> count of itens "below" x.

1) u(x) ≥ u(y) $\implies$ x ≥ y

2) x $\succeq_p$ y $\implies$ u(x) ≥ u(y)

Transitivity: a $\succeq_p$ b $\succeq_p$ c.

Ex: u(c) = 1, u(b) = 2, u(a) = 3.

1) u(x) ≥ u(y) $\implies$ |B(x)| ≥ |B(y)|

By definition: y $\varepsilon$ B(x) x $\succeq_p$ y

y $\nexists$ B(x) |B(y)| - 1 = |B(y) - {y}|

Since |B(x)| ≥ |B(y)|

|B(x)| ≥ |B(y)| - 1

|B(x) - {y}| > |B(y) - {y}|

$\exists$ z $\varepsilon$ X - {y}

x $\succeq_p$ z but y $\succeq_p$ z

Completeness: z $\succeq_p$ y

Transitivity: x $\succeq_p$ y, y $\varepsilon$ B(x) ***-><-**. We started from: |B(x)| ≥ |B(y)|

2) x $\succeq_p$ y $\implies$ u(x) ≥ u(y)

z $\varepsilon$ B(y)

x $\succeq_p$ y; y $\succeq_p$ z

x $\succeq_p$ z $\implies$ z $\varepsilon$ B(x)

B(y) $\subset$ B(x) $\implies$ |B(x)| ≥ |B(y)| -> u(x) ≥ u(y)

Some comments:

1) With finite sets, apparently assuming utility maximization is exactly the same as assuming preferences are complete, reflexive and transitive.

2) Any critique of utility theory is a critique of one of these assumptions (or our notion of preferences). If we are happy with these, then we are happy with utility functions.

3) To turn preferences into *predictive* or *welfare tool*, we need to explicitly state that people will do what they prefer.

##### [**Criticisms of Utility Theory**](https://en.wikipedia.org/wiki/Utility#Discussion_and_criticism):

Some main criticisms of the UT:

1) **Realism**: people don't really act *as-if* they maximize utility, which is not necessarily true: no one chooses objects after consulting some numerical index of goodness. So, what can we **falsify**?

2) **Inconsistent or probabilistic choice**: the individual may choose an object with lower utility for some random reason. But the standard model assumes choices are innate and unchanging. This is the notion of [**random utility**](https://www.sciencedirect.com/topics/social-sciences/random-utility-theory).

3) **Origin of preferences**: tastes change, or [**De Gustibus Non Est Disputandum**](https://www.jstor.org/stable/1807222?seq=1#page_scan_tab_contents).

4) **Procedural vs consequentialist aspects of choice**: If x is chosen from set A, individual is just as well off if given a choice from A vs simply given x without opportunity to choose. But maybe individuals value the freedom to choose, while at the same time too many options can be bad(choice overload).

Despite the criticism, the standard model still remains the workhorse model and remains useful, but important to remember it is **based on assumptions on human behavior** and not laws of nature.

#### Not clear how the content before here is bridged to the next part

**Monotone Definition**: the preference relation $\succeq$ on X is **monotone** if x $\in$ X and y ≥ x implies y $\succeq$ x. If y ≠ x and y ≥ x, then **strictly monotone**. ("goods are goods"???)

[**Locally non-satiated definition**](https://en.wikipedia.org/wiki/Local_nonsatiation): states that for any bundle of goods there is always another bundle of goods arbitrarily close that is preferred to it. Formally, if $\forall$x $\in$ X and all $\delta$ > 0, $\exists$y s.t. ||y - x|| < $\delta$ and y $\succeq$ x. 

Some things to note:

1) Local nonsatiation is implied by **monotonicity of preferences**. Because the converse isn't true, local nonsatiation is a weaker condition.

2) There is no requirement that the preferred bundle y contain more of any good – hence, some goods can be "bads" and preferences can be non-monotone.

3) It rules out the extreme case where all goods are "bads", since the point x = 0 would then be a bliss point.

4) Local nonsatiation can only occur either if the consumption set is unbounded (open) (in other words, it cannot be compact) or if x is on a section of a bounded consumption set sufficiently far away from the ends. Near the ends of a bounded set, there would necessarily be a bliss point where local nonsatiation does not hold.

[**Convex preferences definition**](https://en.wikipedia.org/wiki/Convex_preferences): are an individual's ordering of various outcomes, typically with regard to the amounts of various goods consumed, with the property that, roughly speaking, "averages are better than the extremes". The concept roughly corresponds to the concept of diminishing marginal utility without requiring utility functions. Formally, the preference relation $\succeq$ on X is **convex** if $\forall$x, y $\succeq$ x, z $\succeq$ x and y ≠ z implies that for all $\alpha$ $\in$ (0,1): $\alpha$y + (1 - $\alpha$)z $\succeq$ x. This leads to "moderation in all things".

[**Continuous preference definition**](https://en.wikipedia.org/wiki/Preference_(economics)#Meaning_in_decision_sciences): Continuity simply means that there are **no ‘jumps’** in people's preferences. In mathematical terms, if we prefer point A along a preference curve to point B, points very close to A will also be preferred to B. This allows preference curves to be differentiated. The continuity assumption is "stronger than needed" in the sense that it indeed guarantees the existence of a continuous utility function representation and it is useful because **it does not require the set of preferences to be finite**. Continuity is, therefore, a sufficient condition, but not a necessary one, for a system of preferences. Formally, the preference relation $\succeq$ on X is **continuous** if it is preserved under limits: for any sequence{($x^n,y^n$)}$_{n=1}^{\infty}$ with $x^n$ $\succeq$ $y^n$ for all *n*, x $\succeq$ y, where x and y are limit points of the sequence, respectively.

**The continuity assumption is assumed in order to extend the earlier representation theorem to commodity space X which is not countable.** For continuity proof, see [**here**](https://gabrielvoelcker.netlify.com/mit/debreu_continuityproof.pdf).

[**Debreu's Theorem**](https://en.wikipedia.org/wiki/Debreu_theorems): the Debreu theorems come to answer the following basic question: what conditions on the preference relation of the agent guarantee that such representative utility function can be found? So, his 1954 theorems state that for every preference relation which is complete, transitive and continuous, this relation can be represented by a [**continuous ordinal utility function**](https://en.wikipedia.org/wiki/Ordinal_utility#Continuity). **Slide definition**: supposte the rational preference relation is $\succeq$ on X is continuous. Then there is a continuous utility function u(x) that represents $\succeq$. See p 47 of [**MWG**](https://gabrielvoelcker.netlify.com/mit/Mas-Colell,%20Whinston%20&%20Green%20-%20Microeconomic%20Theory.pdf).

**Choices as primitive**: So far we have taken preferences as a primitive.
Another approach is to take treat choice behavior as the primitive. Let $\mathcal{B}$ be a set of nonempty subsets of X and Ch($\circ$) be a choice rule that assigns a nonempty set of chosen elements Ch(B) $\subset$ B for each B $\in$ $\mathcal{B}$. A lot of modern literature begins this way, more [**here**](https://www.taylorfrancis.com/books/9780429498619).

#### **Recap**

1) Axiomatic method is a useful way to think about complex
assumptions we make.

2) With finite choice sets, the assumption of a utility representation is equivalent requiring preferences to be complete, reflexive, and transitive. Continuity yields utility representation when choice set not finite.

### **Revealed Preferences**

[**Revealed Preferences Theory**](https://en.wikipedia.org/wiki/Revealed_preference):  is a method of analyzing choices made by individuals, mostly used for comparing the influence of policies on consumer behavior. Revealed preference models assume that the preferences of consumers can be revealed by their purchasing habits.

In terms of choice experiment, suppose a consumer is repeatedly asked to choose a consumption bundle x $\in$ $\mathbb{R}^{L+}$. The consumer r is confronted with different prices and we observe resulting choices C = {($p^1,x^1$),($p^2,x^2$),...}. What can we say about preferences and what axioms they satisfy? Need some **structure on preferences**, otherwise indifference could rationalize any behavior.

So, we turn back to the **local nonsatiation condition**: the utility function u: X -> $\mathbb{R}$ satisfies local nonsatiation if $\forall$x $\in$ X, and all $\varepsilon$ > 0, $\exists$y $\in$ X with || y - x || < $\varepsilon$ such that u(y) > u(x). Observe: x,y at price p and p$\circ$x ≥ p$\circ$y. If x chosen when y could have been chosen, x is **directly revealed preferred** to y. The notation for this is **$xR^Dy$**. 

If there is a sequence of preferences such as: $xR^Dz_1$, $z_1R^Dz_2$, ..., $z_nR^Dy$, we can say that x is (indirectly) **revealed preferred** to y and write xRy. Denote strict counterparts with $P^D$ and P, respectively: $xP^Dy$ $\iff$ p$\circ$x > p$\circ$y.

**General Axiom of Revealed Preferences(GARP)**: a set of consumption choices {($p^1,x^1$),($p^2,x^2$),...,($p^n,x^n$)} satisfies the general axiom of revealed preferences if and only if:

$x^iRx^j$ $\implies$ $p^j\circ x^j$ ≤ $p^i\circ x^i$, $\forall i,j$. 

If $x^i$ is revealed preferred to $x^j$, then $x^i$ is only weakly affordable at $x^j$'s prices.

Put differently: if $x^i$ is indirectly revealed preferred to $x^j$, then $x^j$ cannot be strictly directly revealed preferred to $x^i$. That is,

$x^iRx^j$ $\implies$ not $x^jP^Dx^i$

Some textbooks have other axioms of revealed preferences. 

1) [**The Weak Axiom of Revealed Preference(WARP)**](https://en.wikipedia.org/wiki/Revealed_preference#The_Weak_Axiom_of_Revealed_Preference_(WARP)): is one of the criteria which needs to be satisfied in order to make sure that the consumer is consistent with his preferences. If a bundle of goods a is chosen over another bundle b when both are affordable, then the consumer reveals that they prefer a over b. WARP says that when preferences remain the same, there are no circumstances (budget set) where the consumer strictly prefers b over a. By choosing a over b when both bundles are affordable, the consumer reveals that their preferences are such that they will never choose b over a, while prices remain constant. Put simply: if $x^iR^Dx^j$ and $x^i$ ≠ $x^j$, then it is not the case that $x^jR^Dx^i$.

2) [**The Strong Axiom of Revealed Preference(WARP)**](https://en.wikipedia.org/wiki/Revealed_preference#Completeness:_The_Strong_Axiom_of_Revealed_Preferences_(SARP)): The strong axiom of revealed preferences is equivalent to the weak axiom of revealed preferences, except that the consumer is not allowed to be indifferent between the two bundles that are compared. That is, if WARP concludes a $\succeq$ b, SARP goes further and concludes that a $\succ$ b. Put simply: if $x^iRx^j$ and $x^i$ ≠ $x^j$, then it is not the case that $x^jRx^i$.

**Afriat's Proposition**: the central result of consumer theory is Let {($p^1,x^1$),...,($p^n,x^n$)} be a finite set of consumption choices. If a finite set of demand data violates GARP, then these data are inconsistent with choice according to locally nonsatiated, complete, and transitive preferences. Conversely, if a finite set of demand data satisfies GARP, then these data are consistent with choice according to strictly increasing, continuous, convex, complete and transitive preferences.

[**Kreps**](https://gabrielvoelcker.netlify.com/mit/David%20M.%20Kreps-Microeconomic%20Foundations%20I_%20Choice%20and%20Competitive%20Markets-Princeton%20University%20Press%20(2013).pdf): shows the equivalence of GARP and the existence of a solution to a set of linear inequalities. From this  solution, we can construct a utility function rationalizing the choices of the consumer. Chapter 4 of Kreps for complete proof.

- Kreps and Varian emphasize that with finite data on (p, x), strictly increasing or continuous or convex adds no testable restrictions (if satisfy GARP, can construct a utility function satisfying these extra conditions)

GARP is discussed in fields where people are interested in whether some utility function can explain things. Examples: 1) Behavioral: can we explain altruism etc. with utility?; 2) Family/household: are interfamily dynamics causing inefficiency or do families act like they have some common goal?

Then the slides turn to some examples of choices:

1) [**Giving According to GARP: An Experimental Test of the Consistency of Preferences for Altruism**](https://www.jstor.org/stable/2692289?seq=1#page_scan_tab_contents)

2) [**Grandmothers and Granddaughters: Old-Age Pensions and Intrahousehold Allocation in South Africa**](https://www.jstor.org/stable/3990043?seq=1#page_scan_tab_contents)

3) [**Choices, Values, and Frames**](https://www.worldscientific.com/doi/abs/10.1142/9789814417358_0016)

### MWG 1) Preference and Choice

The starting point for any individual decision problem is a set of possible (mutually exclusive) alternatives from which the individual must choose. This set is denoted by X.

There are two approaches to modeling individual choice behavior:

#### 1) *Preference relation*: 
The theory is developed by first imposing rationality axioms on the decision maker's preferences and then analyzing the consequences of these preferences for her choice behavior. The objectives of the decision maker are summarized in a *preference relation*, which we denote by $\succeq$. This is a binary relation on the set of alternatives X, allowing the comparison of pairs of alternatives x,y $\in$ X. x $\succeq$ y can be read as "x is at least as good as y". From $\succeq$ we can derive two other important relations on X:

  a) the **strict preference** relation: $\succ$, defined by x $\succ$ y $\iff$ x $\succeq$ y but not y $\succeq$ x, and it is read as "x is preferred to y".
  
  b) The **indifference** relation, ~, defined by x ~ y $\iff$ x $\succeq$ y and y $\succeq$ x, and read "x is indifferent to y".
  
$\succeq$ is called rational if it possesses the following properties:

- **Completeness**: $\forall$ x,y $\in$ X, we have that x $\succeq$ y or y $\succeq$ x, or both.

- **Transitivity**: $\forall$ x,y,z $\in$ X, if x $\succeq$ y and y $\succeq$ z, then x $\succeq$ z.

Some implications about $\succeq$'s rationality: 

- $\succ$ is both **irreflexive** (x $\succ$ y never holds) and **transitive** (if x $\succ$ y and y $\succ$ z, then x $\succ$ z).

- ~ is **reflexive** (x ~ x for all x), **transitive** (if x ~ y and y ~ z, then x ~ z), and **symmetric** (if x ~ y, then y ~ x). 

- if x $\succ$ y $\succeq$ z, then x $\succ$ z.

#### 2) *Choice rules*: 
Treats the individual's choice behavior as the primitive feature and proceeds by making assumptions directly concerning this behavior. It imposes an element of consistency on choice behavior, paralleling the rationality assumptions of the preference-based approach. Formally, the choice behavior is represented by means of a *choice structure*. A choice structure ($\mathcal{B}$, C($\circ$)) consists of two ingredients: 

a) $\mathcal{B}$ is a family of nonempty sets of X; that is, every element of $\mathcal{B}$ is a set B $\subset$ X. We call the elements B $\in$ $\mathcal{B}$ *budget sets*. The budget sets should be thought of as an exhaustive listing of all the choice experiments that the institutionally, physically or otherwise restricted social situation can conceivably pose to the decision maker.

b) C($\circ$) is a *choice rule* (technically, it is a correspondence) that assigns a nonempty set of chosen elements C(B) $subset$ B for every budget B $\in \mathcal{B}$. The set C(B) may contain one or more than one element. If it contains just one element, that element is the individual's choice from among the alternatives in B. If it contains more than a single element, the elements of C(B) are the alternatives that the decision maker *might* choose, that is, his/her *acceptable alternatives*.

The choice structure ($\mathcal{B}$, C($\circ$)) satisfies the weak axiom of revealed preference if the following property holds:

If x is ever chosen when y is available, then there can be no budget set containing both alternatives for which y is chosen and x is not. If for some B $\in$ $\mathcal{B}$ with x,y $\in$ B we have x $\in$ C(B), then for any B' $\in$ $\mathcal{B}$ with x,y $\in$ B' and y $\in$ C(B'), we must also have x $\in$C(B'). 

A somewhat simpler statement of the weak axiom can be obtained by defining a *revealed preference relation* $\succeq \ast$ from the observed choice behavior in C($\circ$). Given a choice structure ($\mathcal{B}$,C($\circ$)) the revealed preference relation $\succeq \ast$ is defined by x $\succeq \ast$ y $\iff$ there is some B $\in$ $\mathcal{B}$ such that x,y $\in$ B and x $\in$ C(B). We read x $\succeq \ast$ y as "x is revealed at least as good as y". Informally,  we might also say that "x is revealed preferred to y" if there is some B $\in$ $\mathcal{B}$ such that x,y $\in$ B, x $\in$ C(B) and y $\nexists$ C(B), that is, if x is ever chosen over y when both are feasible. Put simply: *if x is revealed at least as good as y, then y can't be revealed preferred to x*.

**Utility functions**: in economics we often describe preference relations by means of a **utility function**. A utilitiy function u(x) assigns a numerical value to each element in X, ranking the elements of X in accordance with the individual's preferences. A function u: X -> $\mathbb{R}$ is a *utility representing preference relation* $\succeq$ if, $\forall$ x,y $\in$ X, x $\succeq$ y $\iff$ u(x) ≥ u(y). A preference relation can only be represented by a function if it is rational.

#### 3) Relationship between Preference Relations and Choice Rules

Consider the following questions: 

**1)** If a decision maker has a rational preference ordering $\succeq$, do her decisions when facing choices from budget sets in $\mathcal{B}$ necessarily generate a choice structure that satisfies the weak axiom?

Yes! Proof: suppose that an individual has a rational preference relation $\succeq$ on X. If this individual faces a nonempty subset of alternatives B $\subset$ X, her preference-maximizing behavior is to choose any one of the elements in the set: C$\ast$(B,$\succeq$) = {x $\in$ B: x $\succeq$ y for every y $\in$ B}

The elements of C$\ast$(B,$\succeq$) are the decision maker's most preferred alternatives in B. In principle, we could have C$\ast$(B, $\succeq$) = $\varnothing$ for some B; but if X is finite, or if suitable(continuity) conditions hold, then C$\ast$(B,$\succeq$) will be nonempty. From now on, we will consider only preferences $\succeq$ and families of budget sets $\mathcal{B}$ such that C$\ast$(B,$\succeq$) is nonempty for all B $\in$ $\mathcal{B}$. We say that the rational preference relation $\succeq$ *generates* the choice structure ($\mathcal{B}$,C$\ast$($\circ,\succeq$)).

Proposition 1.D.1: suppose that $\succeq$ is a rational preference relation. Then the choice structure generated by $\succeq$ ($\mathcal{B}$,C$\ast$($\circ$,$\succeq$)), satisfies the weak axiom. Proof: suppose that for some B $\in$ $\mathcal{B}$, we have x,y $\in$ B and x $\in$C$\ast$(B,$\succeq$). By the definition of C$\ast$(B,$\succeq$), this implies x $\succeq$ y. To check whether the weak axiom holds, suppose that for some B' $\in$ $\mathcal{B}$ with x,y $\in$ B' we have y $\in$ C$\ast$(B',$\succeq$). This implies that y $\succeq$ z $\forall$ z $in$ B'. But we already know that x $\succeq$ y. Hence, by transitivity, x $\succeq$ z for all z $\in$ B', and so x $\in$ C$\ast$(B',$\succeq$). This is precisely the conclusion the weak axiom demands. So, if the behavior is generated by rational preferences, then it satisfies the consistency requirements of the weak axiom.

**2)** If an individual's choice behavior for a family of budget sets $\mathcal{B}$ is captured by a choice structure ($\mathcal{B}$, C($\circ$)) satisfying the weak axiom, is there necessarily a rational preference relation that is consistent with these choices?

Maybe! Given a choice structure ($\mathcal{B},C(\circ$)), we say that the rational preference relation $\succeq$ rationalizes C($\circ$) relative to $\mathcal{B}$ if **C(B) = C$\ast$(B,$\succeq$)** for all B $\in$ $\mathcal{B}$, that is, if $\succeq$ generates the choice structure ($\mathcal{B},C(\circ)$). In words the rational preference $\succeq$ rationalizes the choice rule C($\circ$) on $\mathcal{B}$ if the optimal cnoices generated by $\succeq$(captured by C$\circ$($\circ$,$\succeq$)) coincide with C($\circ$) for all budget sets in $\mathcal{B}$. In a sense, preferences explain behavior; we can interpret the decision maker's choices as if she were a preference maximizer. Note that in general, there may be more than one rationalizing preference relation $\succeq$ for a given choice structure ($\mathcal{B},C(\circ)$).

The proposition above (1.D.1) implies that the weak axiom must be satisfied if there is to be a rationalizing preference relation. In particular, since C$\ast$($\circ$,$\succeq$) satisfies the weak axiom for any $\succeq$, only a choice rule that satisfies the weak axiom can be rationalized. It turns out that the weak axiom is not sufficient to ensure the existence of a rationalizing preference relation.

## **Recitation: Debreu's Theorem**

### 1)  What is Debreu’s theorem? Why should we care?

[**Debreu's Theorems**](https://en.wikipedia.org/wiki/Debreu_theorems): In economics, the Debreu theorems are several statements about the representation of a preference ordering by a real-valued function. Any rational and continuous preference relation $\succeq$ on a (separable,
connected) space X can be represented by a continuous utility
function u : X -> $\mathbb{R}$.

This theorem provides (one) justification for using utility functions
when thinking about choice because: a) it links utility function to appealing and simple axioms; b) Much easier, mathematically, to work with utility functions than preference relations.

#### Proof

Considering the special case:

1) X = $\mathbb{R}_+^n$ for some n $\in$ $\mathbb{N}$.

2) $\succeq$ is strictly monotone.

Some definitions:

1) **Rational**: Complete, transitive and reflexive.

2) **Continuous**: the sets U(x) := {y $\in$ X: y $\succeq$ x} and L(x) := {y $\in$ X : x $\succeq$ y} are closed (contain all their limit points).

3) **Strictly monotone**: y ≥ x $\implies$ y $\succ$ x (actually, we only need y ≥ x $\implies$ y $\succeq$ x and y $\gg$ x $\implies$ y $\succ$ x)

### 2) Overview of proof, and key intuitions

See the proof [**here**](https://gabrielvoelcker.netlify.com/mit/debreu_continuityproof.pdf).

### 3) Details of proof

### 4) Concluding thoughts

## **2) Demand Theory**

### **Demand Theory**

Demand theory is an economic principle relating to the relationship between consumer demand for goods and services and their prices in the market. Demand theory forms the basis for the demand curve, which relates consumer desire to the amount of goods available.


#### [**Demand Theory**](https://en.wikipedia.org/wiki/Supply_and_demand): 
Let **X** $\in$ $\mathbb{R}^{L+}$ be the set of possible consumption vectors in a world with L goods. Let **p** $\in$ $\mathbb{R}^{L++}$ be the price vector. Let $B_{p,w}$ be the **budget set** of a consumer with wealth **w**. 

<center> $B_{p,w}$ = {x $\in$ $\mathbb{R}^{L+}$ | $\sum_{i=1}^n$$p_ix_i$ ≤ w}. </center>

The [**utility maximization problem(UMP)**](https://en.wikipedia.org/wiki/Utility_maximization_problem): is the problem consumers face: "how should I spend my money in order to maximize my utility?" It is a type of optimal decision problem. It consists of choosing how much of each available good or service to consume, taking into account a constraint on total spending as well as the prices of the goods. Formally, it can be written as:

$max_{x\in B_{p,w}}$ u(x) or max$_{x\in \mathbb{R^{L+}}}$u(x) such that p $\circ$ x ≤ w.

Comments:

1) (smallness) consumption decisions do not affect prices.

2) linear prices typically justified (loosely) by possibility of resale.

3) no savings.

#### [**Marshallian Demand/Walrasian Demand Correspondence**](https://en.wikipedia.org/wiki/Marshallian_demand_function): 

Specifies what the consumer would buy in each price and income or wealth situation, assuming it perfectly solves the utility maximization problem. Formally: x(p,w): $\mathbb{R}^{L+}$ x $\mathbb{R}$ $\rightrightarrows$ $\mathbb{R}^{L+}$ is defined by: x(p,w) = {z $\in$ $B_{p,w}$|u(z) = $max_{x \in B_{p,w}}u(x)$}. 

#### **Propositions:**

Suppose u is continuous and satisfies local nonsatiation. Then:

1) The UMP problem has at least one solution.

2) x(p,w) is [**homogenous**](https://en.wikipedia.org/wiki/Homogeneous_function) of degree 0, i.e. x($\alpha$p, $\alpha$w) = x(p,w) for all $\alpha$ > 0.

3) x(p,w) satisfies [**Walras’s law**](https://en.wikipedia.org/wiki/Walras%27s_law)([**2**](https://www.youtube.com/watch?v=5pCeeU5Jbbk)), i.e. p $\circ$ z = w for all z $\in$ x(p,w). An implication of Walras' Law is that the consumers budget will be balanced.

4) If u is strictly quasi-concave, then x(p,w) is a function, i.e.
each x(p,w) contains a single bundle.

#### [**Kuhn-Tucker**](https://en.wikipedia.org/wiki/Karush%E2%80%93Kuhn%E2%80%93Tucker_conditions) necessary conditions:

In mathematical optimization, the Karush–Kuhn–Tucker (KKT) conditions are **first derivative tests** (sometimes called first-order) necessary conditions for a solution in nonlinear programming to be optimal, provided that some regularity conditions are satisfied. Allowing inequality constraints, the KKT approach to nonlinear programming generalizes the method of [**Lagrange multipliers**](https://en.wikipedia.org/wiki/Lagrange_multiplier), which allows only equality constraints. Similar to the Lagrange approach, the constrained maximization (minimization) problem is **rewritten as a Lagrange function whose optimal point is a saddle point**, i.e. a global maximum (minimum) over the domain of the choice variables and a global minimum (maximum) over the multipliers.

So, suppose *u* is [**continuously differentiable**](https://en.wikipedia.org/wiki/Differentiable_function#Differentiability_and_continuity). If $x^{\ast}$ $\in$ x(p,w) is a solution to the UMP, then $\exists$ a constant $\lambda$ ≥ 0 s.t.

<center> $\frac{du}{dx_i} = \lambda p_i \forall i \in {1,2,...,L}$ with $x_i^{\ast} > 0$ </center>

and

<center> $\frac{du}{dx_i} = \lambda p_i$ $\forall i$ $\in$ {1,2,...,L} with $x_i^{\ast}$ = 0 </center>

The constant $\lambda$ is the "Lagrange multiplier".

Suppose $x^{\ast}$ is an [**interior solution**](https://www.thoughtco.com/what-is-an-interior-solution-4082800), and consider feasible perturbation x where agent spends $\Delta$ less on j and $\Delta$ more on i. [**Taylor's theorem**](https://en.wikipedia.org/wiki/Taylor%27s_theorem) implies: 

<center> u(x') $\approx$ u($x^{\ast}$) + $\frac{\partial u}{\partial x_j}$ u($x^{\ast}$) (-$\frac{\Delta}{p_j}$) + $\frac{\partial u}{\partial x_i}$ ($x^{\ast}$) ($\frac{\Delta}{p_i}$) </center>

Since $x^{\ast}$ maximizes utility, perturbation x' can't increase utility; hence, for small $\Delta$,

<center> $\frac{\partial u}{\partial x_j}$($x^{\ast}$)/$p_j$ ≥ $\frac{\partial u}{\partial x_i}$ ($x^{\ast}$) / $p_i$ </center>

Repeat for x'' where spend $\Delta$ more on j, and $\Delta$ less on i. Hence

<center> $\frac{\partial u}{\partial x_j}$($x^{\ast}$) / $p_j$ = $\frac{\partial u}{\partial x_i}$($x^{\ast}$) / $p_i$ </center>

#### Elasticity of Demand

The definition for the **elasticity of demand with income** is 

<center> $\eta_i$ = $\frac{\partial x_i}{\partial w}$ $\circ$ $\frac{w}{x_i}$ = $\frac{\partial log x_i}{\partial log w}$</center>

The definition for the **elasticity of demand with price** is

<center>$\varepsilon_{ij}$ = $\frac{\partial x_i}{\partial p_j}$ $\circ$ $\frac{p_j}{x_i}$ = $\frac{\partial log x_i}{\partial log p_j}$</center>

An appeal of elasticities is that they are unit-less (though it is important to remember they are evaluated at particular points). Elasticities must satisfy “adding up constraints” implied by Walras’
law.

**Proposition**: let $s_i$ be the budget share of the good i, $s_i$ = $\frac{p_ix_i(p,w)}{w}$. Then

1) [**Engel aggregation**](https://www.youtube.com/watch?v=BvXVhd9Y3N8): total expenditure must change by an amount equal to any wealth change(no savings!?). Formally: $\sum_{i=1}^n s_i \eta_i$ = 1. The Engel aggregation means that not all goods are luxuries ($\eta_i >1$), or necessities ($\eta_i <1$) or inferior goods ($\eta_i < 0$).

2) [**Cournot aggregation**](https://www.youtube.com/watch?v=-hkIadrfNGM): $\sum_{i=1}^n s_i \varepsilon_{ij}$ = -$s_j$. The Cournot aggregation says that the budget share $s_j$ of the good j is small when the good j has many substitutes ($\varepsilon_{ij}$ > 0), and big when it has many complements ($\varepsilon_{ij}$ < 0), or their respective budget shares are large.

#### [**Indirect Utility Functions**](https://en.wikipedia.org/wiki/Indirect_utility_function) and [**2**](https://www.youtube.com/watch?v=O3_It-2KEzU)

In economics, a consumer's **indirect utility function** *v(p,w)* gives the consumer's maximal attainable utility when faced with a vector p of goods prices and an amount of income w. It reflects both the consumer's preferences and market conditions.

This function is called indirect because consumers usually think about their preferences in terms of what they consume rather than prices. A consumer's indirect utility v(p, w) can be computed from his or her utility function u(x), defined over vectors x of quantities of consumable goods, by first computing the most preferred affordable bundle, represented by the vector x(p, w) by solving the utility maximization problem, and second, computing the utility u(x(p, w)) the consumer derives from that bundle. The resulting indirect utility function is

<center>**v(p,w) = u(x(p,w))**</center>

##### **Properties**

1) Continuous on $\mathbb{R}_+^n$ x $\mathbb{R}_+$ where *n* is the number of goods;

2) Decreasing in prices;

3) Strictly increasing in income

4) [**Homogenous**](https://en.wikipedia.org/wiki/Homogeneous_function) with degree zero in prices and income; if prices and income are all multiplied by a given constant the same bundle of consumption represents a maximum, so optimal utility does not change;

5) quasi-convex in (p,w).

So, formally the **indirect utility function** v(p,w): $\mathbb{R}^L$ x $\mathbb{R}$ $\rightrightarrows \mathbb{R}$ is defined by: 

v(p,w) = max$_{x \in B_{p,w}}$ u(x)

Alternatively, this is v(p,w) = u(x(p,w)) if x(p,w) is a function or v(p,w) = u($x^{\ast}$) for some $x^{\ast}$ $\in$ x(p,w).

#### [**Theorem of the Maximum**](https://en.wikipedia.org/wiki/Maximum_theorem)

The maximum theorem provides conditions for the continuity of an optimized function and the set of its maximizers with respect to its parameters. The theorem is primarily used in mathematical economics and optimal control.

So, formally: given A $\subset$ $\mathbb{R}^N$ and the closed set Y $\subset$ $\mathbb{R}^K$, the correspondence f: A $\rightrightarrows$ Y is **upper hemicontinuous** if it has a closed graph and images of compact sets are bounded.

- **Closed graph**: for any two sequences $x^m$ -> x and $y^m$ -> y with $x^m$ $\in$ A and $y^m$ $\in$ f($x^m$) $\forall$ m, then y $\in$ f(x).

- Images of compact sets are bounded: for every compact set B $\subset$ A, the set f(B) = {y $\in$ Y : y $\in$ f(x) for some x $\in$ B} is bounded.

##### [**Hemicontinuity**](https://en.wikipedia.org/wiki/Hemicontinuity)

In mathematics, the notion of the continuity of functions is not immediately extensible to **multivalued mappings** or correspondences between two sets A and B. The dual concepts of upper hemicontinuity and lower hemicontinuity facilitate such an extension. Roughly speaking, a function is upper hemicontinuous when a convergent sequence of points in the domain maps to a sequence of sets in the range which contain another convergent sequence, then the image of limiting point in the domain must contain the limit of the sequence in the range. Lower hemicontinuity essentially reverses this, saying if a sequence in the domain converges, given a point in the range of the limit, then you can find a sub-sequence whose image contains a convergent sequence to the given point.

**Lower Hemicontinuity**: A correspondence Γ : A → B is said to be lower hemicontinuous at the point a if for any open set V intersecting Γ(a) there exists a neighbourhood U of a such that Γ(x) intersects V for all x in U. (Here V intersects S means nonempty intersection V $\cap$ S ≠ $\varnothing$.

```{r ,echo=FALSE}
library(png)
library(grid)
setwd("~/Dropbox/R/Site-Gabriel/public/MIT")
img <- readPNG("1.png")
 grid.raster(img)
```

So, given A $\subset$ $\mathbb{R}^n$ and the compact set Y $\subset$ $\mathbb{R}^K$, the correspondence f: A $\rightrightarrows$ Y is **lower hemicontinuous** if $\forall x^m$ -> $x \in A$ with $x^m \in A$ for all *m*, and $y \in f(x)$, we can find a sequence $y^m$ -> y and an integer M such that $y^m \in f(x^m)$ for m > M.

Given the function **f**, we say that it is lower hemicontinuous correspondence iff it is continuous. If a correspondence is both upper and lower hemicontinuous, we say it is continuous.

<center> max *f*: $\mathbb{R}^N$ -> R subject to x $\in$ C(q) </center>

C(q) is a nonempty constraint set and q $\in$ Q belongs to an admissible set Q $\in$ $\mathbb{R}^S$. Suppose *f* is continuous and C(q) is compact for every q $\in$ Q. We already know that there exists a solution. When is the solution “continuous”, when is the objective function “continuous”? Let x(q) be the set of solutions and v(q) be the associated maximum value.

**Theorem**: suppose that f($\circ$) is a continuous function and the constraint correspondence C: Q $\rightrightarrows$ $\mathbb{R}^N$ is continuous. Then the maximizer correspondence x: Q $\rightrightarrows$ $\mathbb{R}^N$ is upper hemicontinuous and the value function v: Q -> $\mathbb{R}$ is continuous.

#### [**Expenditure Minimization Problem**](https://en.wikipedia.org/wiki/Expenditure_minimization_problem)

In microeconomics, the expenditure minimization problem is the dual of the utility maximization problem: "how much money do I need to reach a certain level of happiness?". This question comes in two parts. Given a consumer's utility function, prices, and a utility target,

- how much money would the consumer need? This is answered by the expenditure function.

- what could the consumer buy to meet this utility target while minimizing expenditure? This is answered by the Hicksian demand function.

So, formally: the EMP is:

<center> $min_{x \in \mathbb{R}^{L^+}}$  p $\circ$ x s.t.u(x) ≥ $u_0$ </center>

or

$min_{x \in C_{u_0}}$ p $\circ$ x where $C_{u_0}$ = {z $\in$ $\mathbb{R}_L^+|u(z)$ ≥ $u_0$} 

which is the minimum level of wealth required to achieve utility $u_0$.

**Proposition**: Suppose u(x) is a continuous utility function satisfying local nonsatiation and p $\gg$ 0. Then the solution to the EMP is related to the solution to the UMP by a **basic duality result**:

1) If $x^{\ast}$ is a solution to the UMP with wealth w, then $x^{\ast}$ is a solution to the EMP with utility u($x^{\ast}$).

2) If $x^{\ast}$ is a solution to the EMP with utility $u_0$, then $x^{\ast}$ is a solution to the UMP with wealth p · $x^{\ast}$

The EMP is used mostly to define two functions:

1) [**Expenditure Function**](https://en.wikipedia.org/wiki/Expenditure_function): gives the minimum amount of money an individual needs to spend to achieve some level of utility, given a utility function and the prices of the available goods. So, formally: e: $\mathbb{R}^{L+}$ x $\mathbb{R}$ -> $\mathbb{R}$ is defined by

<center> e(p,u) = $min_{x \in \mathbb{R}^{L+},s.t. u(x)≥u}$ p $\circ$ x or $min_{x \in C_u}$ p $\circ$ x. </center>

2) [**Hicksian Demand Correspondence**](https://en.wikipedia.org/wiki/Hicksian_demand_function): a consumer's Hicksian demand correspondence is the demand of a consumer over a bundle of goods that minimizes their expenditure while delivering a fixed level of utility. If the correspondence is actually a function, it is referred to as the Hicksian demand function, or compensated demand function. So, formally: h: $\mathbb{R}^{L+}$ x $\mathbb{R}$ $\rightrightarrows$ $\mathbb{R}^{L+}$ is defined by

<center> h(p,u) $\equiv$ {z $\in$ $\mathbb{R}^{L+}$|p$\circ$z = e(p,u) and u(z) ≥ u} </center>

or

<center> h(p,u) $\equiv$ arg $min_{z \in C_u}$ p$\circ$z </center>

<center> h(p,u) $\equiv$ {z $\in \mathbb{R}^{L+}|z$ solves EMP at price utility u} </center>

The Hicksian demand is often called the compensated demand function. This is due to the fact that h(p,u) = x(p,e(p,u)), or the Hicksian demand can be used to answer: “how would demand change if we change prices from p to p' 0 **and** give whatever compensation is necessary so the utility level is unchanged.”

**Calculating Hicksian demands**: FOCs for calculating Hicksian demands are: 

<center> $p_i$ = $\lambda \frac{du}{dx_i}$(h(p,u)) if $h_i$(p,u) > 0 </center>

<center> $p_i$ ≥ $\lambda \frac{du}{dx_i}$ (h(p,u)) if $h_i$(p,u) = 0 </center>

Basic properties of the Hicksian demands: suppose u is a continuous utility function satisfying local nonsatiation. Then, the expenditure function e(p, u) is

1) homogenous of degree 1 in p, i.e. e($\alpha$p, u) = $\alpha$e(p, u)

2) trictly increasing in u and non-decreasing in each $p_i$

3) concave in p

4) continuous in p,u

**Properties of Hicksian Demands**: Suppose u is a continuous utility function satisfying local nonsatiation. Then the Hicksian demand correspondence h(p, u) satisfies:

1) h is homogenous of degree 0 in p, i.e. h($\alpha$p,u) = h(p,u), $\forall \alpha,p,u

2) No excess utility property: u(x) = u $\forall$ x $\in$ h(p,u)

3) If u is strictly  quasi-concave, then each h(p, u) contains a
single element, i.e. h(p, u) is a function.

Notes: 

1) is a little different from Walrasian demand. That was homogenous of degree 0 in p and x. Here if we just multiple p and keep u constant, budget is implicitly multiplied.

2) is like Walras’s Law. If x gives extra utility you could have
bought less.

[**Compensated Law of Demand**](https://en.wikipedia.org/wiki/Hicksian_demand_function): One new result we get with Hicksian demand is the Compensated Law of Demand. Suppose *u* is a continuous s, strictly quasiconcave utility function
satisfying local nonsatiation, then for all p' and p'', we have

<center> (p'' - p')$\circ$[h(p'',$u_o$) - h(p',$u_0$)] ≤ 0 </center>

This says that in some weighted average, the demand goes down for products that have gone up in price.

**Corollary**: if $p_i$ increases and all other goods' prices are unchanged, then the Hicksian demand for good i decreases, i.e. $h_i$(p',$u_0$) ≤ $h_i$(p,$u_0$) if $p_i'$ > $p_i$ and $p_j'$ > $p_j$, $\forall j ≠ i$.

**Proposition**: suppose u is a continuous, strictly quasiconcave utility function satisfying local nonsatiation. Then for i = 1, 2, ..., L, we have:

<center> $h_i(p,u)$ = $\frac{d}{dp_i}$e(p,u) </center>

There are a few ways to prove this result. We will use the [**Envelope Theorem**](https://en.wikipedia.org/wiki/Envelope_theorem).

**Corollary**: suppose u is continuous, quasi-concave and satisfies local
nonsatiation and h is continuously differentiable. Then $\frac{\partial h_i}{\partial p_j}$ = $\frac{\partial h_j}{\partial p_i}$

[**Envelope Theorem**](https://en.wikipedia.org/wiki/Envelope_theorem):  is a result about the differentiability properties of the objective function of a parameterized optimization problem. As we change parameters of the objective, the envelope theorem shows that, in a certain sense, changes in the optimizer of the objective do not contribute to the change in the objective function. The envelope theorem is an important tool for comparative statics of optimization models. So, formally, we start with the simple version:

Let $\phi$($\alpha$) = $max_{x \in \mathbb{R}^L}$ f(x;$\alpha$). Suppose $\phi$ is differentiable at $\overline{\rm \alpha}$ and f(x;$\alpha$) is uniquely maximized at x($\overline{\rm \alpha}$). Then

<center> $\frac{d \phi}{d \alpha_i}$ = $\frac{\partial f}{\partial \alpha_i}$(x($\overline{\rm \alpha}$);$\overline{\rm \alpha}$) </center>

**Proof**: follows directly from the chain rule:

<center>$\frac{d \phi}{d \alpha_i}$ = $\frac{\partial f(x(\overline{\rm \alpha});\overline{\rm \alpha})}{\partial \alpha_i}$ + $\sum_{j = 1}^L$ $\frac{\partial f}{\partial x_j}$ (x($\overline{\rm \alpha}$);$\overline{\rm \alpha}$)$\frac{\partial x_j}{\partial \alpha_i}(\overline{\rm \alpha})$ = $\frac{\partial f (x(\overline{\rm \alpha});\overline{\rm \alpha})}{\partial \alpha_i}$ </center>

as the second term drops out since x($\overline{\rm \alpha}$) maximizes f.

##### [**Slutsky Substitution**](https://en.wikipedia.org/wiki/Substitution_effect): 

If income is altered in response to the price change such that a new budget line is drawn passing through the old consumption bundle but with the slope determined by the new prices and the consumer's optimal choice is on this budget line, the resulting change in consumption is called the Slutsky substitution effect. The idea is that the consumer is given enough money to purchase her old bundle at the new prices, and her choice changes are seen.

So, Slutsky thought about compensation a little differently. Instead of just giving enough to achieve the same utility, imagine giving $\Delta$ p $\circ$ x so people can afford same bundle. This usually makes them strictly better off.

**Definition**: the **Slutsky substition matrix** S(p,w) is the L x L matrix with

<center> $s_{ij}$ = $\frac{\partial x_i}{\partial p_j}$ + $\frac{\partial x_i}{\partial w}$ $x_j$ </center>

This means the change in xi resulting from changing $p_j$ **and** giving consumer wealth $x_jdp_j$ to make old bundle affordable.

What happens to the consumption of good j as the price of good i
rises?

1) change in relative price of the k goods (substitution), and

2) reduction of real income of consumer (income effect)

**Proposition**: Suppose u is continuous, locally nonsatiated, and strictly quasi-concave. Let w = e(p, $u_0$). Then

<center> $\frac{\partial h_i}{\partial p_j}$(p,$u_0$) = $\frac{\partial x_i}{\partial p_j}$(p,w) + $\frac{\partial x_i}{\partial w}$(p,w)$x_j$(p,w) = $s_{ij}$ </center>

Rewriting:

<center> $\frac{\partial x_i}{\partial p_j}$(p,w) = $\frac{\partial h_i}{\partial p_j}$(p,$u_0$) - $\frac{\partial x_i}{\partial w}$(p,w)$x_j$(p,w) </center>

The first element after the = is the relative price effect, the second is the real income effect.

**Corollary**: Under the conditions above, the Slutsky substitution matrix is **symmetric** and **negative semi-definite**.

**Symmetry**: can understand why same sign: if rise in price of good i causes demand for good j to rise, then j is in some sense substituting for i. If j substitutes for i, then i must substitute for j. No easy intuition for why they are equal: if 1 cent rise in price of good i means a fall in demand for good j of 10 units, then a 1 cent rise in price of good j means a fall in demand for good i of 10 units.

We saw how h(p, u) is the derivative with respect to p of the EMP’s value function e(p, u). Do not have exact same statement for UMP because Walrasian demand is an ordinal concept.

Do not have exact same statement for UMP because Walrasian demand is an ordinal concept.

[**Roy's Identity**](https://en.wikipedia.org/wiki/Roy%27s_identity): is a major result in microeconomics having applications in consumer choice and the theory of the firm. The lemma relates the ordinary (Marshallian) demand function to the derivatives of the indirect utility function. Specifically, denoting the indirect utility function as v(p,w),} the Marshallian demand function for good i can be calculated as 

<center> $x_i^m$ = $\frac{\frac{\partial v}{\partial p_i}}{\frac{\partial v}{\partial w}} </center>

where p is the price vector of goods and *w* is income. So, it provides analogous result when we normalize utility by marginal utility of wealth.

Suppose that u($\circ$) is continuous utility function and preferences are locally nonsatiated and strictly convex. Suppose that the indirect utility function is differentiable at ($p^{\ast},w^{\ast}$) $\gg$ 0. Then for every i = 1, ..., L,

<center>$x_i(p^{\ast},w^{\ast})$ = -$\frac{\partial v (p^{\ast},w^{\ast})}{\partial p_i}$ / $\frac{\partial v (p^{\ast},w^{\ast})}{\partial w}$ </center>

<center>
```{r ,echo=FALSE}
library(png)
library(grid)
setwd("~/Dropbox/R/Site-Gabriel/public/MIT")
img <- readPNG("2.png")
 grid.raster(img)
```
</center>

## **Useful**{.tabset}

### **Course Info**

Prof: [**Parag Pathak**](https://economics.mit.edu/faculty/ppathak)

[**Syllabus 2018**](https://gabrielvoelcker.netlify.com/mit/syllabus_fall2018.pdf)

Required readings:

- [**Microeconomic Theory - MWG**](https://gabrielvoelcker.netlify.com/mit/Mas-Colell,%20Whinston%20&%20Green%20-%20Microeconomic%20Theory.pdf)

- [**Microeconomic Theory Solutions - MWG**](https://gabrielvoelcker.netlify.com/mit/Mas-Colell, Whinston & Green - Solutions Manual for Microeconomics Theory.pdf)

- [**Microeconomic Foundations I - Kreps**](https://gabrielvoelcker.netlify.com/mit/David%20M.%20Kreps-Microeconomic%20Foundations%20I_%20Choice%20and%20Competitive%20Markets-Princeton%20University%20Press%20(2013).pdf)

### **Problem Sets Solutions**{.tabset}

#### **2017**

[**Problem Set 1**](https://gabrielvoelcker.netlify.com/mit/14.121 pset1_solutions_2017.pdf)

[**Problem Set 2**](https://gabrielvoelcker.netlify.com/mit/14.121 pset2_solutions_2017.pdf)

[**Problem Set 3**](https://gabrielvoelcker.netlify.com/mit/14.121 pset3_solutions_2017.pdf)

[**Problem Set 4**](https://gabrielvoelcker.netlify.com/mit/14.121 pset4_solutions_2017.pdf)

[**Problem Set 5**](https://gabrielvoelcker.netlify.com/mit/14.121 pset5_solutions_2017.pdf)

#### **2018**

[**Problem Set 1**](https://gabrielvoelcker.netlify.com/mit/14.121 ps1sol_2018.pdf)

[**Problem Set 2**](https://gabrielvoelcker.netlify.com/mit/14.121 ps2sol_2018.pdf)

[**Problem Set 3**](https://gabrielvoelcker.netlify.com/mit/14.121 ps3sol_2018.pdf)

[**Problem Set 4**](https://gabrielvoelcker.netlify.com/mit/14.121 ps4sol_2018.pdf)

[**Problem Set 5**](https://gabrielvoelcker.netlify.com/mit/14.121 ps5sol_2018.pdf)

[**Problem Set 6**](https://gabrielvoelcker.netlify.com/mit/14.121 ps6sol_2018.pdf)



















#### **2019**

Gabriel Voelcker 14.121 Microeconomic Theory 1

##### Problem Set 1 

1a) **Transitivity**: it is transitive. $\forall$ x,y,z $\in$ X, let x $\succeq$ y and y $\succeq$ z, then x $\succeq$ z implies transitivity. Given that "≥" is transitive in $\mathbb{R}_+$, we can assume that $\forall$ i, $x_i$ ≥ $y_i$ and $y_i$ ≥ $z_i$ and by transitivity we know that for all *i*, $x_i$ ≥ $z_i$, which implies x $\succeq$ z. 

**Completeness**: it is not complete. Proof by contradiction: given k = 2, let x = {4,2} and y = {2,4}. For i = 1, $x_1$ ≥ $y_1$, but for i = 2, $y_1$ ≥ $x_1$. It concludes that $x_i$ > $y_i$ holds for i = 1 but not for i = 2, which leads to $y_i$ > $x_i$. Since neither x $\succeq$ y nor y $\succeq$ x holds, there is a contradiction and $\succeq$ is not complete.

1b) **Reflexive**: $\succ$ is not reflexive. $\forall$ x $\in$ X, to be reflexive it would need for x $\succeq$ x **and** $\neg$ x $\succeq$ x. A statement and its negation cannot be true together, so $\succ$ is not reflexive.

**Transitive**: $\succ$ is transitive. $\forall$ x,y,z $\in$ X. Let x $\succ$ y and y $\succ$ z. So, $\forall$ i, $x_i$ ≥ $y_i$, $y_i$ ≥ $z_i$. Transitivity will hold if $x_i$ ≥ $y_i$ and $y_i$ ≥ $z_i$, then $x_i$ ≥ $z_i$. Knowing that x ≠ y ≠ z, we have x $\succ$ z.

**Symmetry**: $\succ$ is not symmetric. For $\succ$ to be symmetric, we would need x $\succ$ y and y $\succ$ x to hold. Proof by contradiction: let k = 2 and x = {2,2} and y = {1,1}, we would have that x $\succ$ y but y $\succ$ x doesn't hold, so $\succ$ is not symmetric. 

1c) **Reflexive**: ~ it is reflexive. ~ implies that $x$ ≥ $y$ and $y$ ≥ $x$ $\forall$ x,y. Thus x ~ y means that x = y. By definition, it is clear then that x ~ x.

**Transitive**: it is transitive. Let x ~ y and y ~ z, we have that x = y and y = z. We conclude that x = z and that we have transitivity, because x ~ z.

**Symmetry**: it is symmetric. If x ~ y, we have that x = y, so y ~ x.

2a) C(B) is a frequency distribution over the alternatives of B. We have 6 alternatives in B: 

$B_1$ = {x $\succ$ y $\succ$ z}

$B_2$ = {x $\succ$ z $\succ$ y}

$B_3$ = {y $\succ$ x $\succ$ z}

$B_4$ = {y $\succ$ z $\succ$ x}

$B_5$ = {z $\succ$ x $\succ$ y}

$B_6$ = {z $\succ$ y $\succ$ x}

x is preferred to y in half of the budget sets, and vice versa. The same goes for x and z and y and z. So for a stochastic choice function to be *rationalized by preferences*, we must find a probability distribution Pr over the preference relations on X such that for every budget set, the choice rule is precisely the frequency of choices induced by Pr. By assigning to $\frac{1}{6}$ to each budget set, we have a stochastic choice rationalized by preferences.

2b) On this problem, the probabilities assigned to x $\succ$ y, y $\succ$ z and z $\succ$ x are different, $\frac{1}{4}$. So, the possibility of them all happening together should ≤ $\frac{3}{4}$. However, at least one of the 3 preferences always holds, because if x $\succ$ y and y $\succ$ z don't hold, then y $\succ$ x  and z $\succ$ y, and by transitivity we have that z $\succ$ x, so the stochastic choice function is *not rationalized by preferences* since the sum of all the probabilities cannot exceed 1.

2c) $\alpha$ can gravitate only between $\frac{1}{3}$ and $\frac{2}{3}$. Because of transitivity, one of the three preference relations always holds. So the Pr(x $\succ$ y) or Pr(y $\succ$ z) or Pr(z $\succ$ x) has to equal 1. For this to be true at most, Pr(x $\succ$ y) = Pr(y $\succ$ z) = Pr(z $\succ$ x) = $\frac{1}{3}$.

By assigning a value inferior than $\frac{1}{3}$, for ex: $\frac{1}{5}$, the stochastic function ceases to be rationalized by preferences since the $\neg${x,y}, $\neg${y,z} and {z,x} will have Pr of $\frac{4}{5}$, $\frac{4}{5}$ and $\frac{1}{5}$ respectively. 

### **Exams**{.tabset}

[**2015**](https://gabrielvoelcker.netlify.com/mit/2015exam.pdf)

[**2015 solutions**](https://gabrielvoelcker.netlify.com/mit/2015exam_solutions.pdf)

[**2017**](https://gabrielvoelcker.netlify.com/mit/2017_Final Exam.pdf)

[**2018**](https://gabrielvoelcker.netlify.com/mit/examF18_wSOL_sketch.pdf)
